---
title: "Crossings Annually"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: yeti
---

## *Variations in Border Crossings by Year*

```{r,include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)


knitr::opts_chunk$set(
   fig.width = 12,
   fig.asp = .6,
   fig.height = 12,
   out.width = "90%"
 )

theme_set(theme_minimal() + theme(legend.position = "right"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d


border_crossing_df = readRDS(file = "border_crossing_df_clean.rds")
```


The table below shows us that there has been a decline in border crossings since 2000. There was a steady decline until 2012, where there was a slight upwards turn that carried on through 2018. 

```{r}
## Border crossings by year
border_crossing_df %>% 
  group_by(year) %>% 
  summarize(total_crossings = sum(value)) %>% 
  knitr::kable(col.names = c("Year", "Total Border Crossings")) 

```

Below are a series of plots containing average crossing values across all ports for each month in each of the 5 states with the largest number of border crossings. Each line represents a year from 2000 to 2018. 

Looking at these plots, we can see: 

*   There is a  clear decline in average border crossings over the years across all ports in Michigan and New York. 
*   In Arizona and Texas, there is also an apparent average decline, however, the volume in 2016-2018 shows that there was an upturn in these years compared to years 2012-2016.
*   Lastly, in California, 2018 appears to be the year with the highest average volume across ports, and no clear declines annually. 

```{r}
border_crossing_df %>%
  filter(state %in% c("Texas", "California", "Arizona", "New York", "Michigan")) %>%
  group_by(state, month_name, year) %>%  
  summarize(mean_value = mean(value)) %>%
  ggplot(aes(x = month_name , y = mean_value, color = as.factor(year), group = year)) + 
  geom_point() + 
  geom_line() + 
  theme(legend.position = "bottom") + 
  facet_wrap(~state) +
  labs(title = "Average Number of Border Crossings by Year in New York",
       x = "Month", 
       y = "Number of Crossings", 
       color = "Year") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

 