---
title: "Code for final project"
date: "11/5/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Reading in a tidying the data

Reading in the border crossing data:

*   Time information was removed since all times were 12am and thus were not useful
*   All day information was removed because all were on the 1st of the month for each month
*   Years before 2000 were removed 
*   There is an Eastport (`port_name`) in Maine and Idaho. Create a port_name_state variable instead so there is no confusion 
*   Kept state variable but removed port_name-- we might want to look at state alone
*   I removed 2019 because there is only data for Jan, Feb and March
*   A lot of values were equal to 0. I removed these rows since that means there were no border crossings at that time for those measure types.

```{r clean_data}
month_df <- tibble(month = 1:12, month_name = month.name)

border_crossing_df = read_csv(file = "./Border_Crossing_Entry_Data.csv") %>% 
  janitor::clean_names() %>% 
  mutate(date = str_sub(date, end = -13), 
         location = str_replace(location, "POINT ", ""), 
         port_name_state = str_c(port_name, state, sep = "_")) %>% 
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>% 
    left_join(month_df, by = "month") %>%
  select(-day, -port_name) %>% 
  separate(location, into = c("long", "lat"), sep = " ") %>% 
  mutate(long = str_replace(long, "\\(", ""), 
         lat = str_replace(lat, "\\)", ""), 
         long = as.numeric(long),
         lat =  as.numeric(lat), 
         vehicle = if_else(str_detect(measure, "Passengers") |
                             measure == "Pedestrians",
                           "Non-Vehicles", "Vehicles")) %>% 
  filter(year >= 2000, year < 2019, value != 0)
  
```


```{r save_cleaned_data}
saveRDS(border_crossing_df, file = "border_crossing_df_clean.rds")
```

