---
title: "Border Crossings Report"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    code_folding: hide
---

## Motivations

Immigration has been a hotly contested issue during Trump's presidency. Marking his words of building a "big, beautiful wall" in 2016, President Donald Trump got the green light from The Supreme Court in 2019 and $2.5 billion from the Pentagon's budget to building up the U.S.-Mexico border.

Ironically, despite the promise of his campaign to stop illegal immigration and build a strong bolder, the Trump administration faced a great challenge since the number of apprehensions significantly soar. For example, in August 2018, one-month total apprehensions was the highest on record and an 80 percent increase from July 2018. This trend is quite contrasting to the border crossing situation under the Obama administration. From 2008 to 2016, the yearly apprehensions on the Southwest border declined by 35 percent. 

Although the data we found on border crossings on Kaggle are recorded, legal, and do not run into 2020, we were interested in looking at border crossing trends because of this issue. Understanding the change in legal border crossing is crucial when considering the United States and Mexico's mutually dependent relationship in many aspects: including economy, separated immigrants' family issues, etc. Indeed, Mexico has remained the United States' second-largest export market since 1995, with exports totaling $242 billion in 2017. This analysis will help us understand and show how the border crossings play into life (e.g., movement of goods, commuting, holiday travel).

## Related Work
The following resources are a sample of what inspired this project.

* Trump wall: How much has he actually built?, BBC News, October 2020: [link](https://www.bbc.com/news/world-us-canada-46824649) 

## Initial Questions

* Are there any seasonal trends in border crossings? What are the general trends for vehicle compared to non-vehicle crossings, by month and year?

## Data

### *Source*

Primary data source from Kaggle: [Border Crossing Entry Data](https://www.kaggle.com/akhilv11/border-crossing-entry-data) 

### *Cleaning* 

```{r clean_data, warning=FALSE, message=FALSE}
library(tidyverse)

month_df <- tibble(month = 1:12, month_name = month.name) %>%
         mutate(month_name = as.factor(month_name),
         month_name = ordered(month_name, levels =  c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))) %>%
  arrange(month_name)

border_crossing_df = read_csv(file = "./Border_Crossing_Entry_Data.csv") %>% 
  janitor::clean_names() %>% 
  mutate(date = str_sub(date, end = -13), 
         location = str_replace(location, "POINT ", ""), 
         port_name_state = str_c(port_name, state, sep = "_")) %>% 
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>% 
    left_join(month_df, by = "month") %>%
  select(-day, -port_name) %>% 
  separate(location, into = c("long", "lat"), sep = " ") %>% 
  mutate(long = str_replace(long, "\\(", ""), 
         lat = str_replace(lat, "\\)", ""), 
         long = as.numeric(long),
         lat =  as.numeric(lat), 
         vehicle = if_else(str_detect(measure, "Passengers") |
                             measure == "Pedestrians",
                           "Non-Vehicles", "Vehicles")) %>% 
  filter(year >= 2000, year < 2019, value != 0, state != "Ohio")
  
```


```{r save_cleaned_data, include = FALSE}
saveRDS(border_crossing_df, file = "border_crossing_df_clean.rds")

#save one into dashboard folder as well
saveRDS(border_crossing_df, file = "dashboard/border_crossing_df_clean.rds")

```

After choosing the database, years were limited from 2000-2018, as the information for border crossings in 2019 was only available until March.  Additionally, years prior to 2000 were removed in order to scale the database to a manageable size. Values for Ohio were also removed as observations for this state only began in 2018.    

Information regarding day of the month was removed because the count of all border crossings was reported on the first day of every month of each year. Additionally,all times of the reports were 12am, so time information was removed as well.      

Some ports in different states had the same name (ex: Eastport), thus, a variable `port_name_state` was created in order to make sure each port name was unique. A variable called `vehicle` was also created in order to condense down the wide variety of `measure` entries that were possible.  

Many rows had a `value` of 0 , ie. there were no border crossings for that specific `measure` during that `month` and `year`. These entries were removed as we were only interested in times when border crossings took place.    


#### *Variables of Interest*

**Outcome**
`value`: A count of the number of crossings for each `measure` type at each `port_name_state` in each `month` of each `year`  

**Predictors**

`state`: The state in which the border is located      
`port_code`: The code for the port in which the border entry took place    
`border`: Either the US-Canada or the US-Mexico border    
`month`: Month 1-12 in which the crossing took place    
`month_name`: Month name (Jan-Dec) in which the crossing took place    
`year`: Year from 2000 to 2018 in which the crossing took place    
`measure`: The type of vehicle or non-vehicle utilized in the border crossing. Potential entries include: `r border_crossing_df %>% pull(measure) %>% unique()`    
`long`: Longitude of the port of the crossing  
`lat`: Latitude of the port of the crossing  
`port_name_state`: The port name and state for which the crossing took place. Some ports had the same name so this variable was created to append `state` as well    
`vehicle`: Entry is "non-vehicle" for pedestrians and passengers (on either buses, personal vehicles, or trains), and is "vehicle" for all other `measure` values    

## Exploratory Data Analysis

#### _Crossings Monthly_

**Question: What are the seasonal trends in border crossings?**

Follow this link to [Monthly Crossing Data](crossings_monthly.html).

Graphs in this section visualize border crossings across the months of the year. One visualize produced analyzes trends in total border crossings for both the US-Canada and the US-Mexico border. From this, we were able to see that there is a seasonal variation in border crossings from the US-Canada border, with a jump is crossings during the summer months. However, the US-Mexico border had a stable volume of crossings throughout the months for the years 2000-2018.    

Additionally, border crossings for the top 5 states with the largest total number of border crossings (Arizona, California, Michigan, New York and Texas) were analyzed for 2018. Both the average border crossings across all ports and the total border crossings across all ports for month were evaluated. From this, we found that California had the largest average number of border crossings across all ports, however, Texas had the largest total border crossings by month. This leads us to conclude that Texas has more ports than California, thus the average crossings per port are lower.   

Lastly, from these plots we were able to deduce that New York and Michigan (both on the US-Canada border) had increases in both their total and average crossings by month during the summer months. This trend was not observed in Arizona, California, or Texas, thus supporting the findings that the US-Canada border has an increase in crossings during the summer compared to the stable border crossings on the US-Mexico border.  


#### _Crossings Annually_

**Question: Has the volume of border crossings changed from 2000-2018?**

Follow this link to [Annual Crossing Data](crossings_annually.html).

The analysis in this section contains a table with the total number of border crossings aggregated across all ports for each year. From this, we can see that there has been a decline in border crossings since 2000. The decline was steady decline until 2012 in which a slight upwards turn took place. This upturn took place and began to rise slowly until 2018, although the values never grew larger than they were in 2008.  

Additionally, a graph of average number of border crossings by year for the top 5 states with the largest total number of border crossings (Arizona, California, Michigan, New York and Texas) was created. From this, we can see that for New York and Michigan, there is a clear decline in the average number of border crossings across all ports over the years, with the lowest averages being in 2018. These results contrast to California, in which 2018 saw the highest average of border crossings across all ports. This may lead us to conclude that border crossings on the US-Canada border decreased over time, but maybe the opposite is to be seen for the US-Mexico border.   


#### _Crossings of People_

**Question: What are the trends in border crossings of people through various methods?**

Follow this link to [People Crossing Data](crossings_people.html).

Graphs in this section visualize border crossings over the years at both the US-Canada and US-Mexico border. This visualization was created using only variables that identified they were specifically referencing a person, so that the unit was per person crossing. We can sere were also sharp decines in 2008 at both borders. Train passengers stayed consitent across the two decades at both borders. At the US-Canada border, we saw a fairly consistent number of personal vehicles, with a slight decline over the two decades. Howevever, at the US-Mexico border, we saw a very sharp decline, turning around 2011-2012, with a slight increase from then until 2018.

Pedestrians are nearly non-existant at Canada, which we can assume may be due to the geographic landscape and the temperatures. At the Us-Mexico border, we saw a consistent number of pedestrians that was nearly equivalent to those who use personal vehciles at the US-Canada border, suggesting that there is some equivalency between the two proportions of people and their respective methods of travel.




#### _Crossings of Vehicles_

**Question: What are the general trends for vehicle compared to non-vehicle crossings, by month and year?**

Follow this link to [Crossings of Vehicles](crossings_vehicles.html).

Our graphs visualize both monthly and annual trends of border crossings for vehicles and non-vehicles, which we defined to be any passengers or pedestrians. We decided to explore the natural-log number of crossings due to a very right-skew of the data. This allowed us to examine the differences in trends for vehicle types, and to see that buses and personal vehicles have seasonality at the US-Canada border specifically in the summer months.

We also looked at the overall annual trends in border crossings by vehicles and non-vehicles and saw that they both ebb and flow similarly. This may be because the counts of passengers are correlated with the number of vehicles they correspond to, therefore the differences in the number of crossings is of less importance than the visual trend.

We further stratified our exploration of annual trends by border. We find that visually, there are consistently many more crossings at the US-Mexico border than at the US-Canada border. There is also a bigger gap between the vehicle and non-vehicle lines for the US-Mexico border, likely showing that there are more pedestrians and people per vehicle that for the US-Canada border. The lowest number of crossings at the US-Canada border is in 2009, whereas it is in 2011 at the US-Mexico border.

Finally, we explored the number and type of crossing in Texas and California because a previous chart showed California with a higher average crossing than Texas. Texas has higher total crossings. This visualization is also designed to explore if the general types of vehicles are the same or if they have different patterns overtime. We see stepwise changes in rail crossings and generally the same distributions of crossing types over time. 

## Description of Approaches and Alternate Approaches

This project was inspired by recent US immigration policy and the national discourse about illegal immigration and enforcement. However, the available data only captures legal border crossings. In addition, the dataset does not include information on why a crossing is made, e.g. for immigration, vacation, or commerce. Nor is the nationality of the travel known. 

We adjusted our approach accordingly by coding a "vehicle" and "non-vehicle" variable to understand the movement of people and goods as separately as possible. "Non-vehicle" crossings include passengers and pedestrians.

**Interactive Approach Using Shiny**

We also knew that we wanted to visualize the crossings on an interactive map because the U.S borders are long, especially when including Alaska, making it difficult to see all crossing info at once. Using a Shiny Leaflet map solved this issue by allowing the user to select the US-Canada or US-Mexico border, the year of data, and type of crossing to look at annual trends.  

Users can also see the a boxplot of the monthly LN number of crossings for each year and border to see state variations, due to the log-normal distribution of crossings. Finally, a line graph of LN crossings versus month by vehicle type is included to examine seasonality of vehicle trends. 

## Discussion

In looking at the monthly and yearly trends of border crossings for vehicles and non-vehicles, we saw that the US-Mexico border has a larger volume of crossings and that there is some seasonality in US-Canada border crossings for personal vehicles and buses. We thought we would see seasonality at the US-Canada border during the summer months for summer vacation. We also expected to see a drop in border crossings overall during President Obama's first term around 2008-2012 due to nation-wide immigration issues. We also note that there is less seasonality in all other vehicles and at the US-Mexico border because commercial travel is not influenced by holiday/vacation seasons.

 